A Guide to Pre-Processing High-Throughput Animal Tracking Data

### 2021-02-03 docs/ms_atlas_preproc_main_text_2021-02-03.tex
add updated tex

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.95\textwidth]{figures/fig_02_bounds.png}
    \caption{\textbf{(a)} A velocity-autocorrelated movement track simulated for 5,000 positions (black line) using the \texttt{smoove} package (Gurarie et al. 2017). Three kinds of errors have been artificially added: (1) each position (grey points) is offset from the canonical track with the addition of normally distributed small-scale error, (2) large-scale error has been added to 0.5\% of positions, and (3) 300 positions (indices 500 -- 800) have been displaced to the top-right of the track to simulate a gross distortion that affects a continuous subset of the track. 
    The goal of pre-processing such datasets is to get the estimated positions (grey points) to match the canonical track (black line) as closely as possible.  
    \textbf{(b)} Tracks can be quickly filtered by spatial bounds ($0.5 \leq Y \leq 1.0$; dashed grey lines) using the \texttt{atlastools} function  \texttt{atl\_filter\_bounds}.
    Setting the function argument \texttt{remove\_inside = FALSE} retains positions within user supplied bounds (dashed grey lines; green points), and excludes those outside (grey points).}
    \label{fig:figure_simulated_data}
\end{figure}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.95\textwidth]{figures/fig_03_filter_speed.png}
    \caption{Reducing large-scale outliers in a movement track. The goal is to remove positions far away from the canonical track (black line), and retain those with only reasonable displacements from the canonical track (green points). 
    \textbf{(a)} location error may affect single observations resulting in point outliers or `spikes' (grey circles), but it may also affect continuous subsets of a track, called a `prolonged spike' (purple triangles). 
    The former may be targeted by filtering on appropriate covariates such as speed and turning angle using the \texttt{atlastools} function \texttt{atl\_filter\_covariates}.
    Here, we filter out positions with incoming and outgoing speeds $\geq$ the 90\textsuperscript{th} percentile (0.025) and turning angle $\geq$ the 10\textsuperscript{th} percentile (35$^{\circ}$).
    Howevever, prolonged spikes cannot be effectively corrected by targeting each coordinate pair in isolation.
    \textbf{(b)} The \texttt{atlastools} function \texttt{atl\_remove\_reflections} to remove prolonged spikes (grey circles) in tracking data is an example of targeting track subsets with a common distortion (here, a geometric translation).
    While this method filters out the prolonged spike to return only positions with reasonable displacement from the canonical track (green points) in this example, conceptualising and implementing such algorithms is difficult. 
    Users are cautioned to frequently check this and similar methods' results.}
    \label{fig:figure_filter_bounds}
\end{figure}

@@

% median smooth figure
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.95\textwidth]{figures/fig_04_median_smooth.png}
    \caption{Median smoothing position coordinates reduces small-scale location error in tracking data. 
    \textbf{(a)} The goal of this step is to approximate the simulated canonical track (black line), given positions with small-scale error remaining from previous steps (green points). 
    The canonical track is shown `zoomed in' for better representation, and the missing data at the top of the track (line with no points) is from where the prolonged spike was earlier removed.
    Median smoothing the given position coordinates ((a) green points) over a moving window ($K$) of \textbf{(b)} 3, \textbf{(c)} 5, \textbf{(d)} 11, and \textbf{(e)} 21 positions respectively, yields approximations (purple lines) of the canonical track.
    Users are cautioned that there is no correct $K$, and they must subjectively choose a $K$ which most usefully trades small-scale details of the track for large-scale accuracy.}
    \label{fig:figure_median_smooth}
\end{figure}

@@

% figure for aggregation thinning
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.95\textwidth]{figures/fig_05_thinning.png}
    \caption{Thinning a movement track using aggregation preserves track structure while reducing data volume, but thinning without removing gross location errors and unrealistic movement affects track metrics such as speed.
    (a, b) Movement track positions with an interval of 1 second (green points) aggregated over intervals of (a) 10 and (b) 30 seconds (purple circles), after removing large location errors or filtering unrealistic movement.
    The canonical track is shown as a black line in both panels, while the red line shows the track connecting thinned positions.
    (c) Boxplots show the median and interquartile ranges for speed estimates of tracks aggregated over intervals of 3, 10, 30, and 120 seconds, but without the removal of location errors.
    For comparison, the median and 95\textsuperscript{th} percentile of speed of the canonical track are shown as solid and dashed horizontal lines, respectively.
    Green circles and associated error bars show, in contrast, the median ($\pm$ standard deviation) speed for the same aggregation interval, but after removing large- and small-scale location errors (median smooth $K$ = 11).
    Removing location errors and unrealistic movement before aggregation substantially improves speed estimates, especially at smaller aggregation intervals, as it removes the apparent spikes that contribute to over estimation of speed if not filtered.}
    \label{fig:figure_thinning}
\end{figure}

@@

% first residence patch figure
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.75\textwidth]{figures/fig_06_residence.png}
    \caption{A rotation-advection movement track, similar to the movement of a soaring bird (Gurarie et al. 2017) classified into residence patches with the \texttt{atlastools} function.
    (a) The goal is to segment-cluster the track into areas of prolonged residence (clusters of points) while leaving out the transit between them.
    (b) A plot of residence time against time (solid line; Bracis et al. 2018) shows how the residence patch algorithm segments and clusters positions of prolonged residence. 
    Regions are shaded by the temporal bounds of each residence patch.
    The arguments passed to \texttt{atl\_res\_patch} determine the clustering, and can be adjusted to get a result that fits the study system.
    Users are cautioned that there are no `correct' arguments.
    (c) The residence patch method correctly identifies clusters of positions where the individual is relatively stationary (orange and green patches), as well as positions where it is slowly moving (blue patch).
    This is especially useful when studying more complex behaviour such as area-restricted search, which may have a directional component.
    The function \texttt{atl\_patch\_summary} conveniently presents the classified data for further use, either as an \texttt{sf} object (left: coloured polygons around clustered points), or as summary data of each patch (right: coloured circles).
    While the \texttt{sf} object retains the patch geometry as well as the patch attributes (such as duration), the summary data discards the patch geometry.
    Both spatial and summary objects lose the details of movement between patches (compare black lines of the canonical track [left], and lines connecting patch centroids [right]), but are helpul for questions where the general structure is more relevant.}
    \label{fig:figure_residence_patch}
\end{figure}

@@

% calibration figure
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.75\textwidth]{figures/fig_07_calib_residence_patch.png}
    \caption{Pre-processing steps for WATLAS calibration data showing filtering filtering on speed, median smoothing and thinning by aggregation, and making residence patches.
    \textbf{(a)} Positions with incoming and outgoing speed $\geq$ the 15 m/s are removed (N removed = 13,491, 26.6\%; grey crosses = removed, green points = retained).
    Rectangles show the region expanded in subfigures (b) and (c).
    \textbf{(b)} Expanded view of upper rectangle in (a) showing the raw data (grey crosses), the median smoothed positions (green circles; moving window $K$ = 5), and the smoothed track thinned by aggregation to a 30 second interval (purple squares).
    Square size corresponds to the number of positions used to calculate the averaged position during thinning, with no significant differences.
    Median smoothing retains all 37,324 observations (though changing their coordinates). Thinning aggregates these positions into 1,803 aggregates (4.8\% of the smoothed data volume).
    Median smoothing recovers a good estimate of the true track from significantly over-dispersed raw data, while thinning reduces the data volume but preserves track structure.
    \textbf{(c)} Classifying thinned data into residence patches yields robust estimates of the duration of known stops. The island of Griend (53.25$^{\circ}$N, 5.25$^{\circ}$E) is shown in beige.
    Residence patches (green polygons; function parameters in text) correspond well to the locations of known stops (purple crossed-squares).
    However, the algorithm identified all areas with prolonged residence, including those which were not intended stops, such as stops at the field station (n = 12; green polygon over red crossed-squares).
    The algorithm also failed to find two stops of 6 and 15 seconds duration, since these were lost in the data thinning step (crossed-square without green polygon shows one of these).}
    \label{fig:figure_calibration}
\end{figure}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.75\textwidth]{figures/fig_08_bats.png}
    \caption{Synthesising animal tracks into residence patches can reveal how individuals in a population move in relation to landscape features, prior exploration, and other individuals.
    \textbf{(a)} Linear approximations of the paths (coloured lines) between residence patches (circles) show the movement of three Egyptian fruit bats (\textit{Rousettus aegyptiacus}), tracked over three nights in the Hula Valley, Israel.
    Bat tracks are are shown as thin lines below the linear approximations.
    Colours show bat identity. The grey hexagon represents a roost site.
    Dark green points represent known fruit trees.
    Background is shaded by elevation at 30 metre resolution (SRTM).    
    \textbf{(b)} Spatial representations of an individual bat's residence patches (green polygons) can be used to study site-fidelity by examining overlaps between patches, or to study resource selection by inspecting overlaps with known resources (here, black circles show the location of clusters of fruit trees).
    In addition, the linear approximation of movement between patches (blue paths in panel (a)) can be contrasted with the estimated real path between patches (green lines).
    \textbf{(c)} Residence patch polygons can also show how individuals partition space and resources (black circles are fruit trees), and the combined influence of resource and potential social cues on movement between patches (coloured paths).
    Patches and paths are coloured by bat identity.}
    \label{fig:figure_bats}
\end{figure}

>>>>

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.95\textwidth]{figures/fig_02_filtering_data.png}
    \caption{\textbf{(a)} A velocity-autocorrelated movement track simulated for 5,000 positions (black line) using the \texttt{smoove} package (Gurarie et al. 2017). Three kinds of errors have been artificially added: (1) each position (grey points) is offset from the canonical track with the addition of normally distributed small-scale error, (2) large-scale error has been added to 0.5\% of positions, and (3) 300 positions (indices 500 -- 800) have been displaced to the top-right of the track to simulate a gross distortion that affects a continuous subset of the track. 
    The goal of pre-processing such datasets is to get the estimated positions (grey points) to match the canonical track (black line) as closely as possible.  
    \textbf{(b)} Tracks can be quickly filtered by spatial bounds ($0.5 \leq Y \leq 1.0$; dashed grey lines) using the \texttt{atlastools} function  \texttt{atl\_filter\_bounds}.
    Setting the function argument \texttt{remove\_inside = FALSE} retains positions within user supplied bounds (dashed grey lines; green points), and excludes those outside (grey points).}
    \label{fig:figure_simulated_data}
\end{figure}

@@

% \begin{figure}[h!]
%     \centering
%     \includegraphics[width=0.95\textwidth]{figures/fig_03_filter_speed.png}
%     \caption{Reducing large-scale outliers in a movement track. The goal is to remove positions far away from the canonical track (black line), and retain those with only reasonable displacements from the canonical track (green points). 
%     \textbf{(a)} location error may affect single observations resulting in point outliers or `spikes' (grey circles), but it may also affect continuous subsets of a track, called a `prolonged spike' (purple triangles). 
%     The former may be targeted by filtering on appropriate covariates such as speed and turning angle using the \texttt{atlastools} function \texttt{atl\_filter\_covariates}.
%     Here, we filter out positions with incoming and outgoing speeds $\geq$ the 90\textsuperscript{th} percentile (0.025) and turning angle $\geq$ the 10\textsuperscript{th} percentile (35$^{\circ}$).
%     Howevever, prolonged spikes cannot be effectively corrected by targeting each coordinate pair in isolation.
%     \textbf{(b)} The \texttt{atlastools} function \texttt{atl\_remove\_reflections} to remove prolonged spikes (grey circles) in tracking data is an example of targeting track subsets with a common distortion (here, a geometric translation).
%     While this method filters out the prolonged spike to return only positions with reasonable displacement from the canonical track (green points) in this example, conceptualising and implementing such algorithms is difficult. 
%     Users are cautioned to frequently check this and similar methods' results.}
%     \label{fig:figure_filter_bounds}
% \end{figure}

@@

% median smooth figure
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.95\textwidth]{figures/fig_03_median_smooth.png}
    \caption{Median smoothing position coordinates reduces small-scale location error in tracking data. 
    \textbf{(a)} The goal of this step is to approximate the simulated canonical track (black line), given positions with small-scale error remaining from previous steps (green points). 
    The canonical track is shown `zoomed in' for better representation, and the missing data at the top of the track (line with no points) is from where the prolonged spike was earlier removed.
    Median smoothing the given position coordinates ((a) green points) over a moving window ($K$) of \textbf{(b)} 3, \textbf{(c)} 5, \textbf{(d)} 11, and \textbf{(e)} 21 positions respectively, yields approximations (purple lines) of the canonical track.
    Users are cautioned that there is no correct $K$, and they must subjectively choose a $K$ which most usefully trades small-scale details of the track for large-scale accuracy.}
    \label{fig:figure_median_smooth}
\end{figure}

@@

% figure for aggregation thinning
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.95\textwidth]{figures/fig_04_thinning.png}
    \caption{Thinning a movement track using aggregation preserves track structure while reducing data volume, but thinning without removing gross location errors and unrealistic movement affects track metrics such as speed.
    (a, b) Movement track positions with an interval of 1 second (green points) aggregated over intervals of (a) 10 and (b) 30 seconds (purple circles), after removing large location errors or filtering unrealistic movement.
    The canonical track is shown as a black line in both panels, while the red line shows the track connecting thinned positions.
    (c) Boxplots show the median and interquartile ranges for speed estimates of tracks aggregated over intervals of 3, 10, 30, and 120 seconds, but without the removal of location errors.
    For comparison, the median and 95\textsuperscript{th} percentile of speed of the canonical track are shown as solid and dashed horizontal lines, respectively.
    Green circles and associated error bars show, in contrast, the median ($\pm$ standard deviation) speed for the same aggregation interval, but after removing large- and small-scale location errors (median smooth $K$ = 11).
    Removing location errors and unrealistic movement before aggregation substantially improves speed estimates, especially at smaller aggregation intervals, as it removes the apparent spikes that contribute to over estimation of speed if not filtered.}
    \label{fig:figure_thinning}
\end{figure}

@@

% first residence patch figure
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.75\textwidth]{figures/fig_05_residence.png}
    \caption{A rotation-advection movement track, similar to the movement of a soaring bird (Gurarie et al. 2017) classified into residence patches with the \texttt{atlastools} function.
    (a) The goal is to segment-cluster the track into areas of prolonged residence (clusters of points) while leaving out the transit between them.
    (b) A plot of residence time against time (solid line; Bracis et al. 2018) shows how the residence patch algorithm segments and clusters positions of prolonged residence. 
    Regions are shaded by the temporal bounds of each residence patch.
    The arguments passed to \texttt{atl\_res\_patch} determine the clustering, and can be adjusted to get a result that fits the study system.
    Users are cautioned that there are no `correct' arguments.
    (c) The residence patch method correctly identifies clusters of positions where the individual is relatively stationary (orange and green patches), as well as positions where it is slowly moving (blue patch).
    This is especially useful when studying more complex behaviour such as area-restricted search, which may have a directional component.
    The function \texttt{atl\_patch\_summary} conveniently presents the classified data for further use, either as an \texttt{sf} object (left: coloured polygons around clustered points), or as summary data of each patch (right: coloured circles).
    While the \texttt{sf} object retains the patch geometry as well as the patch attributes (such as duration), the summary data discards the patch geometry.
    Both spatial and summary objects lose the details of movement between patches (compare black lines of the canonical track [left], and lines connecting patch centroids [right]), but are helpul for questions where the general structure is more relevant.}
    \label{fig:figure_residence_patch}
\end{figure}

@@

% calibration figure
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.75\textwidth]{figures/fig_06_calib_residence_patch.png}
    \caption{Pre-processing steps for WATLAS calibration data showing filtering filtering on speed, median smoothing and thinning by aggregation, and making residence patches.
    \textbf{(a)} Positions with incoming and outgoing speed $\geq$ the 15 m/s are removed (N removed = 13,491, 26.6\%; grey crosses = removed, green points = retained).
    Rectangles show the region expanded in subfigures (b) and (c).
    \textbf{(b)} Expanded view of upper rectangle in (a) showing the raw data (grey crosses), the median smoothed positions (green circles; moving window $K$ = 5), and the smoothed track thinned by aggregation to a 30 second interval (purple squares).
    Square size corresponds to the number of positions used to calculate the averaged position during thinning, with no significant differences.
    Median smoothing retains all 37,324 observations (though changing their coordinates). Thinning aggregates these positions into 1,803 aggregates (4.8\% of the smoothed data volume).
    Median smoothing recovers a good estimate of the true track from significantly over-dispersed raw data, while thinning reduces the data volume but preserves track structure.
    \textbf{(c)} Classifying thinned data into residence patches yields robust estimates of the duration of known stops. The island of Griend (53.25$^{\circ}$N, 5.25$^{\circ}$E) is shown in beige.
    Residence patches (green polygons; function parameters in text) correspond well to the locations of known stops (purple crossed-squares).
    However, the algorithm identified all areas with prolonged residence, including those which were not intended stops, such as stops at the field station (n = 12; green polygon over red crossed-squares).
    The algorithm also failed to find two stops of 6 and 15 seconds duration, since these were lost in the data thinning step (crossed-square without green polygon shows one of these).}
    \label{fig:figure_calibration}
\end{figure}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.75\textwidth]{figures/fig_07_bats.png}
    \caption{Synthesising animal tracks into residence patches can reveal how individuals in a population move in relation to landscape features, prior exploration, and other individuals.
    \textbf{(a)} Linear approximations of the paths (coloured lines) between residence patches (circles) show the movement of three Egyptian fruit bats (\textit{Rousettus aegyptiacus}), tracked over three nights in the Hula Valley, Israel.
    Bat tracks are are shown as thin lines below the linear approximations.
    Colours show bat identity. The grey hexagon represents a roost site.
    Dark green points represent known fruit trees.
    Background is shaded by elevation at 30 metre resolution (SRTM).    
    \textbf{(b)} Spatial representations of an individual bat's residence patches (green polygons) can be used to study site-fidelity by examining overlaps between patches, or to study resource selection by inspecting overlaps with known resources (here, black circles show the location of clusters of fruit trees).
    In addition, the linear approximation of movement between patches (blue paths in panel (a)) can be contrasted with the estimated real path between patches (green lines).
    \textbf{(c)} Residence patch polygons can also show how individuals partition space and resources (black circles are fruit trees), and the combined influence of resource and potential social cues on movement between patches (coloured paths).
    Patches and paths are coloured by bat identity.}
    \label{fig:figure_bats}
\end{figure}
